Embedded Domain Specific Languages inside Functional Languages

### 2014-03-13 paper.tex
Basic QC complete

An embedded DSL (EDSL) is a language inside a language.
An EDSL implemented as a library in a host language
that has the look, feel and semantics of its own language,
customized to a specific problem domain.
Haskell~\cite{Haskell98Book}, our functional language of choice, is a great host for EDSLs
because of flexible overloading, a powerful type system, and lazy semantics facilitating this.

As an example of Haskell, consider reversing a list in Haskell
\begin{Code}
reverse :: [a] -> [a]
reverse []     = []
reverse (x:xs) = reverse xs ++ [x]
\end{Code}

This is the complete defintion of reverse: reverse reverses any type of list (the first line);
the reverse of an empty list, written \verb|[]|, is an empty list;
and the reverse of an non-empty list is the reverse of the tail of a linked-list,
appended to the front of the list.

Using this consise syntax, Haskell specifically and the functional programming
community in general have taken the ideas
of EDSLs, which considerably lowered the costs of developing
and maintaining a DSL, and developed many, many of DSL 
which provide higher-level interfaces and abstractions for well-understood systems.

\subsection{Example EDSL: QuickCheck}

As a first example of a EDSL, consider the challange of writing test cases.
Or more specifically, writing the {\em properties\/} that test cases need to satisfy.

\begin{Code}
-- The reverse of a reversed list is itself
prop_reverse (xs :: [Int]) = reverse (reverse xs) == xs
\end{Code}

\verb|prop_reverse| is itself a regular Haskell function,
that takes a list of Int, and returns a boolean, based
on the validity of what is being proposed, specifically
that is a two reverses cancel each other out. 
Here is the neat part -- \verb|prop_reverse| is {\bf also\/} a
domain specific statement, and as such can also be considered
a sub-language inside of Haskell itself.  This style
of using function, in this case function with names
prefixed with \verb|prop_|, taking a number of typed
arguments, and returning a conditional. The property
written in Haskell is also an Embeeded Domain specific
language for properties.

We can run our EDSL, using a function called \verb|quickCheck|:
\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse 
+++ OK, passed 100 tests.
\end{Code}
By running \verb|quickCheck| with our explicit and specific property,
we execute our ESDL inside Haskell. The quickCheck function
generates 100 test cases for our property, and exectutes
them on the fly. If they all hold, then the system prints
a message reflecting this.

In turns out that this sort of mini-language is really useful in practice.

((Go into my QC notes))

>>>>

An EDSL is a library in a host language which has the look, feel and semantics of its own language,
customized to a specific problem domain.
Using Haskell's concise syntax, Haskell specifically and the functional programming
community in general have taken the ideas
of EDSLs, which considerably lowered the costs of developing
and maintaining a DSL, and developed many, many of DSL 
which provide higher-level interfaces and abstractions for well-understood systems.

We look at two examples of DSLs, one for automatically generating test cases for software testing,
and a second for specifying hardware circuit behaviors.

\subsection{Example EDSL: QuickCheck Properties}

As a first example of a EDSL, consider the challenge of writing test cases.
Or more specifically, writing the {\em properties\/} that test cases need to satisfy.

\begin{Code}
-- The reverse of a reverse'd list is itself
prop_reverse_twice (xs :: [Int]) = reverse (reverse xs) == xs
\end{Code}

\verb|prop_reverse_twice| is a regular Haskell function,
that takes a list of Int, and returns a boolean, based
on the validity of what is being proposed, specifically
that is a two reverses cancel each other out. 
Here is the neat part -- \verb|prop_reverse_twice| is {\bf also\/} a
domain specific statement, and as such can also be considered
a sub-language inside of Haskell itself.  This style
of using functions, in this case function with names
prefixed with \verb|prop_|, taking a number of typed
arguments, and returning a conditional is a small language.
The property written in Haskell is also an Embedded Domain Specific
Language for properties.

We can run our EDSL, using a function called \verb|quickCheck|:
\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse_twice
+++ OK, passed 100 tests.
\end{Code}
By running \verb|quickCheck| with our explicit and specific property,
we execute our ESDL inside Haskell. The quickCheck function
generates 100 test cases for our property, and executes
them on the fly. If they all hold, then the system prints
a message reflecting this. The test cases are generated using
the type class system -- QuickCheck gives specific types the
power of test-case generation -- and the \verb|quickCheck| function
uses this to generate random tests.

Properties are useless unless they are falsify-able. As an examples
of an incorrect property, consider this property for reverse.
\begin{Code}
prop_reverse xs ys = reverse xs ++ reverse ys == reverse (xs ++ ys)
\end{Code}
This is stating that the reverse of two distinct lists
is the same as the reverse of both lists appended together.
But this property is false.
\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse 
FALSE!!!
\end{Code}

In turns out that this sort of mini-language is really useful in practice.
Despite the simplicity of how Haskell is being used, the QuickCheck EDSL
provided a way of thinking about and directly expressing properties.
In the full QuickCheck EDSL has additional functionality, including
the ability to generate random function arguments, the ability to control
the distribution of the random test cases, and the ability to state pre-conditions
to a property. From this language, several other implementations of
these idea have been constructed. There is even a swedish company, ???,
which sells a QuickCheck for Erlang, a language for large scale concurrency
as found in telephone switches, and the language used by snapchat, the recent
Facebook \$19 billion dollar purchase.
